SparseIDS: Learning Packet Sampling with Reinforcement Learning

### 2020-01-26 paper_rl/paper.tex
Added some text to the continuous section

@@

However, for practical use, not only classification accuracy and security are important but also computational efficiency. This follows from the fact that a rationally-acting company will only deploy an \gls{ids} if the cost that is saved by avoiding security breaches is higher than the cost of the \gls{ids} itself. The cost of the \gls{ids} can be split in the cost of purchase/installation, the cost of operating it and the cost of maintaining it. The goal we pursue here is to develop sampling strategies that do not significantly degrade classification performance while being able to only process a small fraction of the original data. Specifically, the sampling technique should
\begin{enumerate}
\item \textbf{choose optimally}, taking only the data that contain the most information and skip the less relevant ones. 
\item have a parameter that allows to \textbf{trade off} classification performance for sparsity.
\item be \textbf{independent of the classifier} so that any classifier can be used and that the classifier doesn't have to be aware of the sampling strategy. 
\item be \textbf{retrainable} so that the sampling can be continuously adapted depending on the current threat landscape. 
\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not and the remaining packets can be skipped altogether. 
\end{enumerate}

@@

As \ours{} uses a parameter to trade off accuracy for sparsity, the resulting accuracy and sparsity are determined by this parameter. However, since it also might be desired to specify the sparsity itself, we propose a steering system which takes a desired sparsity as its input and modifies the parameter of the reinforcement learning in a control loop until the specified sparsity is reached.

@@

\todo{Max: Add figure}
The classification reward metric captures how correct the classifier is on average for all future packets of a flow. The label of each packet in a flow has a value of 0 for non-attack traffic and 1 for attack traffic with one flow being only attack or non-attack but not mixed. The confidence is the sigmoided ($\frac{e^x}{1+e^x}$) output of the classifier, yielding a value between 0 and 1, where 0 is absolute confidence for a flow being benign and 1 is absolute confidence for a flow being an attack. 
\begin{align}
\begin{split}
R_{\text{classification},t} = \frac{1}{T-t} \sum_{i=t+1}^{T} |\text{label}_t - \text{confidence}_t|
\end{split}
\end{align}

@@

Combining these objectives, the actor network aims to minimize the loss function
\begin{align}
\begin{split}
l_{\text{a},t} =& -\log \left( \pi \left( a_t \given s_t ; \theta_\text{a} \right) \right) U_t \\ 
&- \beta H\left( \pi\left( s_t; \theta_a \right)\right)
\end{split}
\end{align}

@@

@@

\subsection{Continuous Actions}
\todo{Max: Add formulas for the Log-Normal distribution}

>>>>

% Fares:
%As the need for security mechanisms in today's infrastructure communication networks grows, a substantial body of scientific work focuses on developing reliable, explainable frameworks for the identification of unwanted behaviors. In the past, a viable way to detect intrusion was to analyze the contents of the packets themselves and determine, for example, whether a packet contains potentially harmful content by matching patterns. More recently, with the increasing deployment of encryption, the focus is now on features that are always available for network monitoring devices such as port numbers, protocol flags, packet sizes when encrypting above the transport layer.

@@

% Fares:
%A network communication is typically aggregated into \textit{flows}, which are commonly defined as a sequence of packets that share certain properties. When analyzing flows, not only the aforementioned features are available but also features representing the temporal behavior of the individual packets. Various approaches have been proposed to extract flows features after which anomaly detection can be performed, in \cite{meghdouri_analysis_2018} Meghdouri et al. compare such approaches.
%While these approaches often work well, it is troublesome because, in the worst case, the whole flow must be received first and only then anomaly detection can be applied; revealing malicious flows. Thus, we design a network \gls{ids} that operates on a per-packet basis and determines whether packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
%At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
%In \cite{hartl_explainability_2019}, Hartl et al. show that such an architecture has similar performance to other flow-based anomaly detection systems, but can still detect anomalies \textit{before} the flow ends and analyzed its robustness with respect to adversaries extensively.

However, for practical use, not only classification accuracy and security are important but also computational efficiency. This follows from the fact that a rationally-acting company will only deploy an \gls{ids} if the cost that is saved by avoiding security breaches is higher than the cost of the \gls{ids} itself. The cost of the \gls{ids} can be split in the cost of purchase/installation, the cost of operating it and the cost of maintaining it. The goal we pursue here is to develop sampling strategies that do not significantly degrade classification performance while being able to only process a small fraction of the original data. Specifically, the sampling technique should

% Fares:
%Nevertheless, not only the accuracy and reliability of classification, but also the computational efficiency, are critical for practical use. This follows from the fact that a rationally-acting organization will only implement an \gls{ids} if the cost saved by preventing security breaches is higher than the cost of the \gls{ids} itself. Noteworthy that the cost of an \gls{ids} can be divided into (a) the cost of purchase/installation, (b) the cost of operation and (c) the cost of maintenance.  The goal we pursue here is to develop autonomous sampling strategies that do not significantly degrade classification performance while being able to only process a small fraction of the original data. Specifically, the sampling technique should:

\begin{enumerate}
\item \textbf{choose optimally}, taking only the data that contain the most information and skip the less relevant ones. 
% Fares:
%\item \textbf{choose optimally}, taking only samples that contain the most information and skip the less relevant ones. 
\item have a parameter that allows to \textbf{trade off} classification performance for sparsity.
\item be \textbf{independent of the classifier} so that any classifier can be used and that the classifier doesn't have to be aware of the sampling strategy. 
\item be \textbf{retrainable} so that the sampling can be continuously adapted depending on the current threat landscape. 
\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not and the remaining packets can be skipped altogether. 
\end{enumerate}

@@

% Fares:
%For this purpose we develop a \gls{rl}-based \gls{ids}: \textit{\ours{}} that fulfills the above properties. We show that a significant number of packets can be skipped while the accuracy does not drop
%considerably. Compared to other common sampling techniques, \ours{} performs better when being trained on the same number of packets.

As \ours{} uses a parameter to trade-off accuracy for sparsity, the resulting accuracy and sparsity are determined by this parameter. However, since it also might be desired to specify the sparsity itself, we propose a steering system which takes a desired sparsity as its input and modifies the parameter of the reinforcement learning in a control loop until the specified sparsity is reached.

% Fares:
%The accuracy-sparsity trade-off of \ours{} can be regulated by a reinforcement learning parameter that directly affects both later quantities. However, in certain situations, a certain sparsity is to be achieved, hence we further develop and propose a steering system that fluctuates such a trade-off parameter in order to obtain a coveted sparsity in a closed control loop.

@@

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{img/rewards.pdf}
  \caption{The computation of the rewards for a sample flow: Gray packets are skipped while white ones are chosen. At the top row the resulting rewards for classification for each chosen packet are depicted and in the bottom row the ones for sparsity. Numbers are rounded to two digits after the decimal point.}
  \label{fig:rewards}
\end{figure}

The classification reward metric captures how correct the classifier is on average for all future packets of a flow. The label of each packet in a flow has a value of 0 for non-attack traffic and 1 for attack traffic with one flow being only attack or non-attack but not mixed. The confidence is the sigmoided ($\frac{e^x}{1+e^x}$) output of the classifier, yielding a value between 0 and 1, where 0 is absolute confidence for a flow being benign and 1 is absolute confidence for a flow being an attack. 
\begin{align}
\begin{split}
R_{\text{classification},t} = \frac{1}{T-t} \sum_{i=t+1}^{T} |\text{label}_t - \text{confidence}_t|
\end{split}
\end{align}

@@

Combining these objectives, the actor network aims to minimize the loss function
\begin{align}
\begin{split}
l_{\text{a},t} =& -\log \left( \pi \left( a_t \given s_t ; \theta_\text{a} \right) \right) U_t \\ 
&- \beta H\left( \pi\left( s_t; \theta_a \right)\right)
\end{split}\label{eq:actor}
\end{align}

@@

An overview of the neural networks of the classifier, actor and critic that we use in this work is given in \autoref{fig:neuralNetworkArchitecture}.

@@

A schematic overview over the \gls{rl} for sampling is given in \autoref{fig:sampling}.

\subsection{Continuous Actions}

For performing continious actions we need a probability distribution which can be parametrized to have the mean between 0 and positive infinity. Furthermore it must have at least two parameters so that the standard deviation, which is required for the entropy (\autoref{eq:actor}), can be changed independent of the mean. For example, the Exponential distribution can be parametrized to have its mean anywhere between 0 and positive infinity, however, it has only one parameter and thus the mean and the standard deviation cannot be independently changed. Conversely, a simple distribution which fits our requirements is the Log-normal distribution, which is defined as a normal distribution whose output is exponentiated:

Let $Z$ be standard normally distributed and $\mu_{\text{normal}}$ and $\sigma_{\text{normal}} > 0$ be two real numbers. Then 
\begin{align}
X = e^{\mu_{\text{normal}}+\sigma_{\text{normal}} Z}
\end{align}
is log-normally distributed.

The actor network (\autoref{fig:neuralNetworkArchitecture}) shows that the actor network has two outputs: $\mu$ and $\sigma$. These encode the mean and standard deviation of a log-normal distribution. Since the mean and the standard deviation of a log-normal distribution can never be smaller or equal to zero, we always apply the softplus function to them, so that the actor can only output $\mu$ and $\sigma$ that are larger than 0: 
\begin{align}
\text{softplus}(x) = \log\left(1 + e^x \right)
\end{align}
To parametrize the normal distribution that underlies the log-normal distribution, we first have to compute the parameters $\mu_{\text{normal}}$ and $\sigma_{\text{normal}}$ from $\mu$ and $\sigma$. We can compute them as follows:

\begin{align}
\mu_{\text{normal}} = \frac{1}{2}\left( 2\log(\mu) - \log\left(\frac{\mu^2 + \sigma^2}{\mu^2}\right) \right)
\end{align}

\begin{align}
\sigma_{\text{normal}} = \sqrt{\log\left(\frac{\mu^2 + \sigma^2}{\mu^2}\right)}
\end{align}
